An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-22 paper.tex
Begin to restructure introduction

Let's imagine a game played by three people
$A$, $B$, and $C$. Suppose each of them has a piece of information,
say $X$, $Y$, and $Z$ respectively. The goal of the game is
to get information from other people as much as possible.
Suppose $A$ talks to $B$ first and then $B$ talks to $C$,
then $C$ may has all three pieces of information $X$, $Y$, and $Z$ even
if $C$ has never talked to $A$.
However, if $B$ talks to $C$ first and then $A$ talks to $B$, then
no matter how smart $C$ is, $C$ can never get information
$X$.  Time is an important factor for information flow.

We can analyze the spread of information between $A$, $B$, and $C$
as a process on a graph that changes over time.
Such graphs are known as evolving graphs.
% Evolving graphs are models to study this kind of scenario where the
% interrelations between entities are time-dependent.
% We assume that the reader is familiar with the basic graph terminology.
% All the necessary material can be found in}.
We follow the notation of (static) graphs in \cite{even12,kegi11}.
In general, an evolving graph $G(t) = (V(t), E(t))$ is  a graph where the node set
and edge set are functions of time $t$.
However, in different literature the term ``evolving graph'' has been
used to refer to slightly different models.
 Flajolet et al. \cite{fkp89} use the term to mean the evolution of random graphs.
Bahmani et al. \cite{bkmu12} use the term to mean a sequence of graphs where
the difference between two consecutive graphs is small.
Similar models are also investigated using the term ``evolving network'' and
``time-dependent graph''.

In this work, we consider a discretization of the time line as $n$ distinct
time stamps and define an evolving graph $G_n$ to be
a sequence of (static) graphs
$G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$.
Depends on applications, the graph $G^{[i]}$ could mean either
a snapshot of $G(t)$ at time stamp $i$ or  the union of all the graphs
between time stamp $i-1$ and $i$ (excludes time stamp $i-1$).
Both the node set and edge set of $G^{[i]}$ could be different from
these of $G^{[j]}, j \ne i$.
\begin{center}
\begin{tikzpicture}[snake=zigzag, line before snake = 5mm, line after snake = 5mm]
    % draw horizontal line
    \draw (0,0) -- (2,0);
    \draw[snake] (2,0) -- (4,0);
    \draw (4,0) -- (5,0);
    \draw[snake] (5,0) -- (7,0);

@@

    % draw nodes
    \draw (0,0) node[below=3pt] {$ 1 $} node[above=3pt] {$   $};
    \draw (1,0) node[below=3pt] {$ 2 $} node[above=3pt] {$ G^{[2]} $};
    \draw (2,0) node[below=3pt] {$ 3 $} node[above=3pt] {$  $};
    \draw (3,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (4,0) node[below=3pt] {$ 5 $} node[above=3pt] {$  $};
    \draw (5,0) node[below=3pt] {$ 6 $} node[above=3pt] {$ G^{[6]}$};
    \draw (6,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (7,0) node[below=3pt] {$ n $} node[above=3pt] {$ $};
  \end{tikzpicture}
\end{center}
Our model is the same as that of Tang et al. \cite{tmml10} though they
use slightly different notation.
Grindrod et al.'s model  \cite{gphe11} \cite{grihig13} is very similar to ours,
where an evolving graph is considered as a sequence of adjacency matrices.
Our model can also be seen as a special case of
the multi-layer networks discussed by Kivel{\"a} et al. \cite{kabg14}, where
there is only one set of layers for time.
% We present a preliminary discussion of using the model to mine citation networks.

@@

We show a small evolving graph $ \langle G^{[1]}, G^{[2]}, G^{[3]}\rangle$ with $3$ time stamps $t_1$, $t_2$ and $t_3$
in Figure \ref{fig:eg_shortest_path}, where each $G^{[i]}$ is a graph at time stamp $t_i$.
For example,
at time stamp $t_1$, the graph $G^{[1]}$ contains three nodes $1$, $2$ and $3$ and $(1,2)$ is an edge of $G^{[1]}$.
Though Figure \ref{fig:eg_shortest_path} presents a
directed evolving graph,  the algorithms that we shall discuss work on both
directed and undirected evolving graphs.
We shall use this example throughout the paper to illustrate various
notions.

@@

In order to define the evolving-graph BFS, we need the concept of a
\emph{temporal path}.
A temporal path on an evolving graph $G_n$ is a sequence of node-time stamp
pair $\langle (v_1, t_1), (v_2, t_2), \ldots, (v_m, t_m) \rangle$ such that
$t_1 \leq t_2 \leq \cdots \leq t_m$ and $v_i = v_j$ if and only if
$t_i \ne t_j$. The length of a temporal path is the length of the sequence.
For example, $\langle (1, t_1), (1, t_2), (3, t_2), (3, t_3) \rangle$ and
$\langle (1, t_1), (2, t_1), (2, t_3), (3, t_3) \rangle$
are two temporal paths from $(1, t_1)$ to $(3, t_3)$
on the evolving graph in Figure \ref{fig:eg_shortest_path}.

@@

  \foreach \from/\to in {n1/n4}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=70, in=40] (n11);
  \draw[dotted,->](n10) -- (n7);
   \draw[dotted,->](n11) -- (n12);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{An evolving graph with 3 time stamps $t_1$, $t_2$ and $t_3$.
The back dashed lines of the two figures indicate two temporal paths from $(1, t_1)$ to $(3,t_3)$.}
\label{fig:active}
\end{figure}

>>>>

Let's imagine a game played by three people, numbered $1$, $2$, and $3$,
each of whom has a message, labeled $a$, $b$, and $c$ respectively.
At each turn, one particular player is allowed to talk to one other player,
who must in turn convey all the messages in his or her possession.
The goal of the game is to collect all the messages.
Suppose $1$ talks to $2$ first, and $2$ in turn talks to $3$.
Then, $3$ can collect all the messages even without talking to $1$ directly.
However, if $2$ talks to $3$ before $1$ talks to $2$, then
$3$ can never get $a$.

We can use analyze the spread of information between the players using graph
theory. In this process, the time ordering of events matters, which leads us to
consider a graph that changes over time, i.e.\ a graph $G(t) = (V(t), E(t))$ whose
vertex set $V$ and edge set $E$ evolve over on time, $t$.
Such a graph is called an ``evolving graph''~\cite{fkp89,bkmu12},
``evolving network'' or %XXX WHO?
``time-dependent graph''. %XXX WHO?

Treatments of evolving graphs vary in their generality and focus.
Kivel{\"a} \textit{et al.}~\cite{kabg14} treats time dependence as a special
case of families of graphs with multiple interrelationships.
Others like Flajolet \textit{et al.}~\cite{fkp89} use time to index the family
of related graph, but are not concerned with explicit time-dependent processes.

Our model can also be seen as a special case of
the multi-layer networks discussed by Kivel{\"a} \textit{et al.}~\cite{kabg14}, where
there is only one set of layers for time.

%XXX is the first sentence of this restatement true?
Previous work on evolving graphs tends to focus on special cases, such as
random graphs~\cite{fkp89} or sequences of graphs whose successive differences
are small~\cite{bkmu12}.  use the term to mean the evolution of random graphs.
Bahmani et al. \cite{bkmu12} use the term to mean a sequence of graphs where
successive differences are small.

Here, we are interested in extensions of graph algorithms that generalize for evolving graphs that work in full
generality.

%ARE THIS RELATED TO THE PREVIOUS WORK HOW
Work in this area include Despite the fact that graph metrics \cite{ntmm13,tmml09,tsmm09} and
various centrality measures like Katz centrality~\cite{gphe11,grihig13},
betweenness \cite{alhi15}, and closeness \cite{tmml10})
have been generalized for evolving graphs,
a BFS for evolving graphs, which we shall call it \emph{evolving-graph BFS},
was never explicitly described or verified.
%XXX THIS LEADS US TO CONSIDER ALGEBRAIC FORMULATIONS ALSO SINCE...

In this paper, we present a generalization of the breadth first search (BFS)
algorithm to evolving graphs. BFS is arguably the most fundamental graph search
algorithm, which leads to other graphical computations
such as the minimum spanning tree and shortest path \cite{ckrs09}.
We also demonstrate how BFS on evolving graphs can be expressed in the language
of linear algebra, and how our formulation helps analyze citation networks.

\subsection{Temporal Paths over Active Nodes}

The key new idea in generalizing BFS to evolving graphs is to be able to compute
traversals that span multiple times. We call some, but not all,
such traversals \textit{temporal paths}.

Figure~\ref{fig:eg_shortest_path} shows a small example of a evolving
directed graph, $G_3 = \langle G^{[1]}, G^{[2]}, G^{[3]}\rangle$, consisting of
a sequence of three graphs $G^{[i]}$, each bearing a time stamp $t_i$.
There are directed edges $1\rightarrow2$ at time $t_1$,
$1\rightarrow3$ at time $t_2$, and $2\rightarrow3$ at time $t_3$.
Each edge exists only at a particular discrete time and the nodes connected
by edges are considered \textit{active} at that time.

Temporal paths connect only active nodes in ways that respect time ordering.
Thus the sequences
$\langle (1, t_1), (1, t_2), (3, t_2), (3, t_3) \rangle$ and
$\langle (1, t_1), (2, t_1), (2, t_3), (3, t_3) \rangle$
are both examples of \textit{temporal paths} from $(1, t_1)$ to $(3, t_3)$,
which are drawn as dotted lines with arrowheads in Figure~\ref{fig:active}.
However, $\langle (1, t_1), (1, t_2), (2, t_2), (3, t_2), (3, t_3) \rangle$
is not a temporal path because node 2 is inactive at time $t_2$.
Similarly,

  \foreach \from/\to in {n1/n4}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=70, in=40] (n11);
  \draw[dotted,->](n10) -- (n7);
   \draw[dotted,->](n11) -- (n12);
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(4,3) -- (-2,3) node [midway,yshift=-0.6cm]{ $t_2$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(-4,3) -- (-10.5,3) node [midway,yshift=-0.6cm]{ $t_1$};
\draw [decorate,decoration={brace,amplitude=5pt},xshift=-4pt,yshift=0pt]
(11.5,3) -- (5.5,3) node [midway,yshift=-0.6cm]{ $t_3$};
    \end{tikzpicture}
\end{center}
\caption{An evolving graph with 3 time stamps $t_1$, $t_2$ and $t_3$.
The black dashed lines of the two figures indicate two temporal paths from $(1, t_1)$ to $(3,t_3)$.}
\label{fig:active}
\end{figure}

The restriction that temporal paths may only traverse active nodes reflects
underlying structure in many real world applications, such as analyzing influence
of nodes over social networks, and furthermore leads to interesting and nontrivial
adaptations of graph algorithms such as BFS

This example shows a directed
evolving graph
For example,
at time stamp $t_1$, the graph $G^{[1]}$ contains three nodes $1$, $2$ and $3$ and $(1,2)$ is an edge of $G^{[1]}$.
Though Figure \ref{fig:eg_shortest_path} presents a
directed evolving graph,  the algorithms that we shall discuss work on both
directed and undirected evolving graphs.
We shall use this example throughout the paper to illustrate various
notions.

We follow the notation of (static) graphs in \cite{even12,kegi11}.
In this work, we consider a discretization of the time line as $n$ distinct
time stamps and define an evolving graph $G_n$ to be
a sequence of (static) graphs
$G_n = \langle G^{[1]}, G^{[2]}, \ldots, G^{[n]} \rangle$.
Depends on applications, the graph $G^{[i]}$ could mean either
a snapshot of $G(t)$ at time stamp $i$ or  the union of all the graphs
between time stamp $i-1$ and $i$ (excludes time stamp $i-1$).
Both the node set and edge set of $G^{[i]}$ could be different from
these of $G^{[j]}, j \ne i$.
\begin{center}
\begin{tikzpicture}[snake=zigzag, line before snake = 5mm, line after snake = 5mm]
    % draw horizontal line
    \draw (0,0) -- (2,0);
    \draw[snake] (2,0) -- (4,0);
    \draw (4,0) -- (5,0);
    \draw[snake] (5,0) -- (7,0);

@@

    % draw nodes
    \draw (0,0) node[below=3pt] {$ 1 $} node[above=3pt] {$   $};
    \draw (1,0) node[below=3pt] {$ 2 $} node[above=3pt] {$ G^{[2]} $};
    \draw (2,0) node[below=3pt] {$ 3 $} node[above=3pt] {$  $};
    \draw (3,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (4,0) node[below=3pt] {$ 5 $} node[above=3pt] {$  $};
    \draw (5,0) node[below=3pt] {$ 6 $} node[above=3pt] {$ G^{[6]}$};
    \draw (6,0) node[below=3pt] {$  $} node[above=3pt] {$  $};
    \draw (7,0) node[below=3pt] {$ n $} node[above=3pt] {$ $};
  \end{tikzpicture}
\end{center}
Our model is the same as that of Tang et al. \cite{tmml10} though they
use slightly different notation.
Grindrod et al.'s model  \cite{gphe11} \cite{grihig13} is very similar to ours,
where an evolving graph is considered as a sequence of adjacency matrices.
% We present a preliminary discussion of using the model to mine citation networks.

@@

@@

In order to define the evolving-graph BFS, we need the concept of a
\emph{temporal path}.
A temporal path on an evolving graph $G_n$ is a sequence of node-time stamp
pair $\langle (v_1, t_1), (v_2, t_2), \ldots, (v_m, t_m) \rangle$ such that
$t_1 \leq t_2 \leq \cdots \leq t_m$ and $v_i = v_j$ if and only if
$t_i \ne t_j$. The length of a temporal path is the length of the sequence.

@@
